---
output: github_document
---


<!-- README.md is generated from README.Rmd. Please edit that file -->


```{r setup, echo = FALSE, message = FALSE}
options(width = 100)
Sys.setlocale("LC_COLLATE", "en_US.UTF-8") # ensure common sorting envir
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
ver <- desc::desc_get_version(".")
ver <- paste0("https://img.shields.io/badge/Version-", ver,
              "-success.svg?style=flat&logo=github")
sample_data <- splyr:::sample.adat
```


# The `splyr` package

<!-- badges: start -->
![GitHub version](`r ver`)
[![CRAN status](http://www.r-pkg.org/badges/version/splyr)](https://cran.r-project.org/package=splyr)
[![R-CMD-check](https://github.com/stufield/splyr/workflows/R-CMD-check/badge.svg)](https://github.com/stufield/splyr/actions)
[![](https://cranlogs.r-pkg.org/badges/grand-total/splyr)](https://cran.r-project.org/package=splyr)
[![Codecov test coverage](https://codecov.io/gh/stufield/splyr/branch/main/graph/badge.svg)](https://app.codecov.io/gh/stufield/splyr?branch=main)
[![Lifecycle: stable](https://img.shields.io/badge/lifecycle-stable-brightgreen.svg)](https://lifecycle.r-lib.org/articles/stages.html#stable)
[![License: MIT](https://img.shields.io/badge/License-MIT-blue.svg)](https://choosealicense.com/licenses/mit/)
<!-- badges: end -->


## Overview

The `splyr` package contains the general functions necessary to manipulate
internal `R` representations of `*.adat` files and wrangle those data into
convenient form for analysis.


----------------


## Installation

```{r install-git, eval = FALSE}
remotes::install_github("splyr")
```


----------------


## Usage

To load `splyr` simply make a call to `library()` as usual:

```{r load, message = FALSE}
library(splyr)
```


## Help summary of the package

```{r help, eval = FALSE}
library(help = splyr)
```


-------------


## Useful functions in `splyr`


### Transforming Data

* `center_scale()`
* `create_recipe()`

```{r center-scale}
scaled <- center_scale(sim_adat)
apply(feature_matrix(scaled), 2, mean) |> sum()  # mean = 0
apply(feature_matrix(scaled), 2, sd)             # sd = 1

# `create_recipe()`
rcp <- create_recipe(sim_adat)
rcp
```

-----------


```{r helpr, echo = FALSE}
library(helpr)   # get_outliers
```

### Imputing Data

* `impute_outliers()`, `get_outliers()` (`helpr`)
* `imputeNAs()`
* `impute_predictors()`

```{r imputing}
# Outliers
x <- withr::with_seed(1, rnorm(10))   # normal
x <- c(x, 100)                        # add outlier
get_outliers(x)                       # index of the outlier

# parameters stored in attributes (parametric only)
attributes(get_outliers(x, type = "para"))

# Impute 11th value
impute_outliers(x)

# Impute NAs
x <- withr::with_seed(1, rnorm(6))
x[ c(3, 5) ] <- NA
median(x, na.rm = TRUE)

imputeNAs(x)

table(imputeNAs(x))

# Predictors
x   <- data.frame(a = 1:3, b = 4:6, c = 7:9, d = c(1.23, 4.56, 7.89))
tbl <- tibble::tribble(
  ~ Feature,  ~ xtrm_max, ~ impute_max, ~ xtrm_min, ~ impute_min,
    "a",         NA,        NA,           NA,         NA,
    "b",         5,         5,            0,          1,
    "c",         9,         7,            7.1,        7.1
)
impute_predictors(x, tbl)
```

-----------


### Binding Data

* `bind_intersect()`
* `bind_union()`

```{r bind}
df1 <- data.frame(a = 1, b = 2, c = 3, row.names = "A")
df2 <- data.frame(a = 4, b = 5, d = 6, row.names = "B")
df3 <- data.frame(a = 7, b = 8, e = 9, row.names = "C")
list_df  <- list(a = df1, b = df2, c = df3)
list_df

bind_intersect(list_df)

bind_union(list_df)
```

-----------

### Selecting Feature Data

* `refactor_data()`
* `feature_matrix()`

```{r refactor}
df  <- data.frame(a = factor(c("a", "b")), b = 1:2L)
foo <- df[df$a == "a", ]
foo

levels(foo$a)   # 2 levels! "b" is a ghost level

bar <- refactor_data(foo)
levels(bar$a)   # 1 level now
```

-----------

### `seq.XXXX` formats

* `seq_lookup()`
* `seqify()`
* `lookup_anno()`

```{r seq-lookup}
seqs <- withr::with_seed(101, sample(names(sample_data), 10L))
seqs

# NAs for those analytes dropped from menu
seq_lookup(seqs)

seqify(seqs)

# Pass `tbl` containing annotations
# to reconstitute those missing ones
seq_lookup(seqs, splyr:::apt_data)

# OR: on the fly
seq_lookup(seqs, attr(sample_data, "Col.Meta"))
```

