---
title: "Safely Rename Data Frames"
author: "Stu Field"
description: >
  How to safely and reliably rename variable names of a
  data frame in R.
output:
  rmarkdown::html_vignette:
    fig_caption: yes
vignette: >
  %\VignetteIndexEntry{Safely Rename Data Frames}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
library(wranglr)
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```


# Introduction

Renaming variables/features of a data frame (or `tibble`) 
is a common task in data science. Doing so *safely* is often a struggle.
This can be achieved *safely* via the `dplyr::rename()` function via 2
steps:

1. Set up the mapping in either a named vector
1. Apply the `dplyr::rename()` function via `!!!` syntax
1. Alternatively, roll-your-own `rename()` function

* **Note**: all entries in the mapping (i.e. key) object *must* be
  present as `names` in the data frame object.


## Example with `mtcars`

```{r rename-df}
# Create map/key of the names to map
key <- c(MPG = "mpg", CARB = "carb")   # named vector
key

# rename `mtcars`
rename(mtcars, !!! key) |> head()
```

------------------


## Alternative to `dplyr`
If you prefer to avoid the `dplyr` import/dependency, you can achieve a
similar result with similar syntax by writing your own renaming function:

```{r rename2}
rename2 <- function (.data, ...) {
  map <- c(...)
  loc <- setNames(match(map, names(.data), nomatch = 0L), names(map))
  loc <- loc[loc > 0L]
  newnames <- names(.data)
  newnames[loc] <- names(loc)
  setNames(.data, newnames)
}
```

Now, with *similar* syntax (but cannot use `!!!`):

```{r rename-usage}
# rename `mtcars` in-line
rename2(mtcars, MPG = "mpg", CARB = "carb") |>
  head()

# rename `mtcars` via named `key`
rename2(mtcars, key) |>
  head()
```

